### 2020-06-15 chapters/background.tex
Proposed approaches before proposed method
Aggregation properties and families

In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

@@

For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and for other step values as well.

@@

As it turns out, besides \textit{invertibility}, there are other properties that allow us to group aggregations in different families. Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

\textcolor{red}{TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation}

\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/aggregation-operations-properties.png}
      \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

Based on the \textit{combine} and \textit{lower} functions, the following are formal definitions of these properties. An aggregation is said to have:

@@

    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

@@

%We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}. Numerical variables can be measured. When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}. On the other hand, Strings are sequences of characters. The \textit{string} type is the most generic data type in that any other type can be converted to a string. Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3". Similarly, a date is a sequence of characters and thus a string. However, despite being a string, it has semantics unlike a regular one. Knowing the date format, one can extract the day, month and year. For that reason we make the distinction between the \textit{string} and \textit{date} data types. Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values. A variable is assigned to a particular category based on its properties. The categorical variable is also a sequence of characters and therefore a string variable. The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.

@@

\section{Exponential Moving Averages}

>>>>

In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not truly exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

@@

For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families. Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{query} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties. An aggregation is said to have:

@@

    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

\begin{table}[!htb]
\begin{center}
     \begin{tabular}{p{2cm}|p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.6cm}|}
                 & \rotatebox{35}{\textbf{invertible combine}} & \rotatebox{35}{\textbf{associative combine}} & \rotatebox{35}{\textbf{commutative combine}} & \rotatebox{35}{\textbf{size-preserving combine}} & \rotatebox{35}{\textbf{unary query}}
                 \\
                 \hline
    \textbf{sum-like}     & \checkmark & \checkmark               & \checkmark               & \checkmark                   & \checkmark       \\
    \textbf{collect-like} & \checkmark              & \checkmark               & $\times$               & $\times$                   & ?           \\
    \textbf{median-like}  & \checkmark              & \checkmark               & \checkmark               & $\times$                   & ?           \\
    \textbf{max-like}     & $\times$              & \checkmark               & ?                   & \checkmark                   & \checkmark       \\
    \textbf{sketch-like}  & $\times$              & \checkmark               & \checkmark               & \checkmark                   & $\times$      
    \end{tabular}
    \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
  \label{tbl:aggregations-properties}
    \end{center}
\end{table}

According to these properties, Tangwongsan et al. define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. These are the sum, collect, median, max and sketch-like families. The sum-like family of aggregations comprises aggregations that have an invertible, associative, commutative and size-preserving \textit{combine} as well as a unary \textit{query} function. Examples of such aggregations are the sum, count, mean/average and standard deviation operations. The collect-like family of aggregations has an invertible and associative \textit{combine} function but their \textit{query} function is not always unary. For example, keeping a string concatenation aggregation is invertible and associative but not commutative nor size-preserving. While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.} the 5th youngest element). Hence, their \textit{query} function is not always unary. The median-like family of aggregations contains for example the percentile operation. For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.} percentile 95 or 99). The max-like family comprises aggregations like the maximum or minimum, operations that are not invertible but associative and size-preserving, as well as unary from a \textit{query} function perspective. The final family, the sketch-like family, comprises aggregations that are usually solved by approximate aggregators, also called probabilistic data structures or \textit{sketches}. These aggregations can be membership queries, item frequency estimation or cardinality estimation.

@@

%We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}. Numerical variables can be measured. When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}. On the other hand, Strings are sequences of characters. The \textit{string} type is the most generic data type in that any other type can be converted to a string. Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3". Similarly, a date is a sequence of characters and thus a string. However, despite being a string, it has semantics, unlike a regular one. Knowing the date format, one can extract the day, month and year. For that reason we make the distinction between the \textit{string} and \textit{date} data types. Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values. A variable is assigned to a particular category based on its properties. The categorical variable is also a sequence of characters and therefore a string variable. The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.

@@

\section{Exponential Moving Averages} \label{sec:emas}

### 2020-06-15 chapters/introduction.tex
Aggregation properties and families

The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relativily to a reference period.

@@

\textcolor{red}{TODO how to define success criteria without "reverse engineering" this?}

The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

@@

In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.

In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.

>>>>

The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.

@@

The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item low memory footprint

    \item low latency to work in real-time

    \item fix a reference period as the normal state of the system

    \item explainable alerts
\end{itemize}

@@

%In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.

%In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.

### 2020-06-15 chapters/my-work.tex
Aggregation properties and families

Our goal is to detect data pattern shifts in an online fashion. Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase. Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window. Doing this would mean our memory consumption grew linearly with window or reference period size. We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes. In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.65]{figures/approach2-timeseries.png}
      \caption[]{Alert state for feature \textit{x} in window 2}
      \label{fig:approach2-timeseries}
    \end{center}
\end{figure}

@@

\section{Proposed method: Numerical Feature Monitoring}
%what we want (reiterate)
%A reference aggregation snapshot to compare with a streaming aggregation that must be kept in real-time and constant in memory usage

%- introduce method as two-phased method (batch and streaming)

% only for numerical features for now

\subsection{Stream monitoring goals}
%- what is our goal in the streaming phase?
           % - alert when streaming fields/features/variables' distribution changes considerably
          %  - need to choose an aggregation that allows us to know the distribution of variables -> histogram for each feature
               % - lightweight -> approximated EMA-based histogram (recall EMAs background), each bin count is an EMA-count aggregation
               % - describe an EMA approximated histogram and how it discounts old events, the simulated window size, ...
           % - we can compute an exact histogram over the reference period because it is done in batch, where resources like RAM or time are not heavily constrained in constrast to streaming systems
           % - given 2 histograms (ref hist and stream hist) we can apply distance metrics (such as a JSD)
             %   - worth mentioning other metrics can be used (Wasserstein distance, Kolmogorov–Smirnov, ...)
         %   - but how to threshold this "distance" value? ----> we need to know the distribution of "distance" values
             %   - how is this accomplished
             %   - agressive sampling within ref period of smaller periods (equal in size to the stream simulated window)
              %  - give a practical use-case where it makes a diference:
                 %   - left-skewed distro vs right-skewed distro ---> threshold is very different (95 or 99 percentile for example)

>>>>

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/approach2-timeseries.png}
      \caption[]{Two different time-series with same max, min, mean and standard deviation}
      \label{fig:approach2-timeseries}
    \end{center}
\end{figure}

@@

\section{Proposed method: Numerical Feature Monitoring}

Our goal is to detect data pattern shifts in an online fashion. Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase. Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window. Doing this would require storing all window contents hence making our memory consumption grow linearly with sliding window or reference period size. We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes. In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.

To that end, we devise a two-phased method. The first phase relies on a batch analysis that builds our aggregated reference snapshot to be later on compared with the sliding one. The second phase focus on efficiently updating the sliding window aggregation snapshot and occasionally perform a similarity test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types as well, as we will discuss.

\subsection{Stream monitoring goals}
Before diving into the batch and streaming algorithms we must further define our goal. In this Section we pose the question: \textit{"what is our goal during the streaming or monitoring phase?"} and answer it.

A data stream is an infinite stream of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are also named features or time-dependent variables. You can think of them as columns in a traditional relational database table. Given this context, our goal is to produce alerts of the following form:

\[\textit{timestamp: [(x1, 0.23), (x2, 0.51), (x3, 0.55)}]\]

The format used allows us to know the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (for instance, a float value between 0 and 1 that works as a confidence value).

An alert should be raised if a feature distribution changes considerably. This immediately poses challenges such as \textit{(a)} finding a suitable aggregation that encodes each feature's distribution and allows for a change detection test and \textit{(b)} defining a deviation threshold $\alpha$.

\subsection*{Approximated Histogram Aggregation}

We chose to encode a feature distribution as a histogram. A histogram is an aggregation that represents the distribution of numerical features. However, we must recall that we need to keep this histogram in constant time in a sliding window fashion: that implies we need to be able to evict old elements and insert new ones.

An exact histogram aggregation must keep the entire sliding window in memory...

\textcolor{red}{TODO: review this section after EMAs is written}

\subsection*{Defining the deviation threshold}

Assume we use a distance function that takes two histograms, our reference aggregation histogram and our sliding window aggregation histogram, and outputs a distance or difference value. How do we threshold this value? For instance, consider the following scenario, where ...

\textcolor{red}{TODO: give an example of left and right skewed distros}
%  - give a practical use-case where it makes a diference:
 %   - left-skewed distro vs right-skewed distro ---> threshold is very different (95 or 99 percentile for example)

Hard-coding a threshold value to be used for all features is cumbersome, impractical and error prone. The distance threshold must be computed based on the distribution of distances. How do we get to know the distribution of such distances? That is one of the goals for the batch phase.

### 2020-06-15 chapters/sota.tex
Aggregation properties and families

\textcolor{red}{TODO: explain this better, paper not clear to me}

@@

This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions. The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function.

\textcolor{red}{\textbf{TODO: develop this idea, they talk about Kolmogorov–Smirnov and others, maybe relate to model monitoring paper, talk about JSD, Wasserstein, so on so forth}}

@@

\textcolor{red}{TODO: Add table with SWAG algorithms, time and space complexity and applicable restrictions on the aggregation}

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection*{False Positive rate}
% error bound related with parameters
The false positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}. It is assumed that the hash functions produce an independent and random index value for each element. Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.

@@

\subsection*{Time complexity analysis}
When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell. Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}. Similarly, querying the frequency of an element \textit{el} consists in computing the minimum of all row values. Since both insertion and query have constant time complexity, the CMS is constant in time.

@@

The authors claim that the proposed sliding version remains as accurate as the original HLL aggregator. In the context of their work, each element in the set is denoted as a "flow" and they state that \textit{"In practice, with an additional memory of only 35kB, a standard error of about 3\% can be achieved for a data stream of several million flows"}.

>>>>

@@

This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions. The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function. They provide as examples of a possible \textit{d} function implementation the  Kolmogorov–Smirnov Test \cite{Kolmogorov-Smirnov} and the Wilcoxon-Signed-Rank Test \cite{Wilcoxon}.

@@

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection*{False Positive rate}
% error bound related with parameters
The false-positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}. It is assumed that the hash functions produce an independent and random index value for each element. Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.

@@

\subsection*{Time complexity analysis}
When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell. Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}. Similarly, querying the frequency of an element \textit{el} consists of computing the minimum of all row values. Since both insertion and query have constant time complexity, the CMS is constant in time.

@@

The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator. In the context of their work, each element in the set is denoted as a "flow" and they state that \textit{"In practice, with an additional memory of only 35kB, a standard error of about 3\% can be achieved for a data stream of several million flows"}.
